---
title: "Projected losses of global mammal and bird ecological strategies"
author: "Robert Cooke"
date: "08/01/2019"
output: html_notebook
---

#### --------------------------------------------------------------
## Step 0: setup ##
#### --------------------------------------------------------------

#### Set up required packages
```{r}
if (!require("pacman")) install.packages("pacman")
pacman::p_load(dplyr, plyr, readr, tibble, FD, ade4, cowplot, mice, reshape2, tidyr, ks, hypervolume, alphahull, purrr, TTR, plotrix, agricolae, psych)

# dplyr: used for data manipulation # calls: arrange, bind_rows, mutate, mutate_at, %>%, select, rename, filter, left_join
# plyr: used for data manipulation
# readr: used to read csvs # calls: read_csv
# tibble: used to convert column to rownames and vice versa # calls: column_to_rownames, rownames_to_column
# FD: calculates gower dissimilarity # calls: gowdis
# ade4: principal coordinate analysis, permutation tests # calls: dudi.pco, as.randtest
# cowplot: plotting # calls: ggplot
# mice: multiple imputation with chained equations # calls: md.pattern, mice, complete
# reshape2: data manipulation
# tidyr: data manipulation # calls: nest, unnest
# ks: kernel density estimation # calls: Hpi, kde
# hypervolume: used to calculate hypervolumes # calls: hypervolume_svm, hypervolume_set, hypervolume_overlap_statistics, plot.HypervolumeList
# alphahull: needed for contour plotting # called in plot.HypervolumeList
# purrr: data manipulation # calls: map2
# TTR: cumulative sd # calls: runSD
# plotrix: se # calls: std.error
# agricolae: kruskal-wallis test # calls: kruskal
# psych: data manipulation # calls: decribeBy
```

#### --------------------------------------------------------------
## Step 1: load ##
#### --------------------------------------------------------------

#### Load data

```{r}
# load: trait data
trait <- readr::read_csv("data/trait.csv")

# load: extinction risk data
iucn_ex <- readRDS("data/df_iucn_ex.rds")

# load: class data
bi_class <- readRDS("data/df_bi_class.rds")

```

#### --------------------------------------------------------------
## Step 2: diet PCoA ##
#### --------------------------------------------------------------

#### Calculate synthetic diet trait (principal component from a PCoA of 10 diet categories)

```{r}
diet_all <- trait %>% 
  # select diet data
  dplyr::select(binomial, contains("diet")) %>% 
  # drop species with missing data: 1178
  dplyr::filter(!is.na(diet_inv)) %>%   
  # add species names to rownames (needed for gowdis function)
  tibble::column_to_rownames(var = "binomial") %>% 
  as.data.frame()

# calculate species x species gower distance matrix based on traits
diet_gd <- FD::gowdis(diet_all)

# perform principal coordinates analysis (PCoA)
diet_pco <- ade4::dudi.pco(diet_gd, scannf = FALSE)

pc_diet <- diet_pco$tab

summary(diet_pco)
# Projected inertia Ax1 = 36.180
# Projected inertia Ax2 = 15.729

# principle component axes
pcomp_diet <- as.data.frame(diet_pco$tab[,1:2]) %>% 
  tibble::rownames_to_column(var = "binomial")

# diet category projection
n <- nrow(diet_all)
points_stand <- scale(diet_pco$tab[,1:2])
S <- cov(diet_all, points_stand)
U <- S %*% diag((diet_pco$eig[1:2]/(n-1))^(-0.5))
colnames(U) <- colnames(diet_pco$tab[,1:2])

# diet categoires (see Wilman et al., 2014)
U <- as.data.frame(U) %>% 
  mutate(trait = c("Inv", "Vend", "Vect", "Vfish", "Vunk", "Scav", "Fruit", "Nect", "Seed", "Planto"))
# Inv - Invertebrates # Vend - Vertebrate endotherms # Vect - Vertebrate ectotherms # Vfish - Fish # Vunk - Vertebrate unknown or general # Scav - Scavenge # Fruit - Fruit # Nect - Nectar # Seed - Seeds # Planto - Other plant material

# scale diet category arrows
mult <- min(
  (max(pcomp_diet$A2) - min(pcomp_diet$A2)/(max(U$A2)-min(U$A2))),
  (max(pcomp_diet$A1) - min(pcomp_diet$A1)/(max(U$A1)-min(U$A1)))
)

U <- U %>% 
  mutate(v1 = 0.0003 * mult * A1) %>% 
  mutate(v2 = 0.0003 * mult * A2)

# plot diet PCoA
pcoa_diet <- ggplot(pcomp_diet, aes(A1, A2)) +
  # set up plot
  geom_hline(yintercept = 0, size = 0.2, lty = 2, colour = "grey") + 
  geom_vline(xintercept = 0, size = 0.2, lty = 2, colour = "grey") +
  # add origin lines
  #geom_text(alpha = 0.4, size = 1, aes(label = binomial))
  geom_point() +
  # add species
  coord_equal() +
  geom_segment(data = U, aes(x = 0, y = 0, xend = v1, yend = v2), arrow = arrow(length = unit(0.2, "cm")), colour = "darkgrey") +
  # add arrows
  geom_text(data = U, aes(x = v1, y = v2, label = trait), size = 4, colour = "darkgrey",
            nudge_y = c(rep(0, 6), 0.005, 0.005, 0.0005, -0.004), 
            nudge_x = c(0.005, rep(0, 7), -0.009, 0)) +
  # add arrow labels
  labs(x = "PC1 (36.2%)", y = "PC2 (15.7%)") +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text = element_text(colour = "black"),
        legend.position = "none",
        text = element_text(size = 20))

pcoa_diet
```

#### --------------------------------------------------------------
## Step 3: trait imputation ##
#### --------------------------------------------------------------

#### Impute missing trait data using the traits themselves and phylogenetic eigenvectors

#### Set up trait data (add diet PCoA axis)

```{r}
tr <- dplyr::select(trait, -contains("diet")) %>% 
  # add PCoA diet data
  dplyr::left_join(pcomp_diet, by = "binomial") %>% 
  # add class data
  dplyr::left_join(bi_class, by = "binomial") %>% 
  # reorder variables + drop pc_diet2
  dplyr::select(everything(), diet_pc1 = A1, class, -A2)
```

#### Need to do imputation separately for mammals and birds, as we are using separate trees

#### Mammal trait imputation

```{r}
tr_mam <- filter(tr, class == "Mammalia") %>% 
  dplyr::select(-class)

## phylogenetic data

# load: t_pem_mam 
t_pem_mam <- readRDS("data/df_t_pem_mam.rds")
# phylogenetic eigenvectors matched to species names - using mammal supertree from Fritz et al., 2009 "Geographical variation in predictors of mammalian extinction risk: big is bad, but only in the tropics"

# join trait and phylogenetic data (use first ten eigenvectors - Penone et al., 2014)
tr_mam <- dplyr::left_join(tr_mam, t_pem_mam, by = "binomial")

# missing data pattern
mice::md.pattern(dplyr::select(tr_mam, -binomial, -c(V_1:V_10)))

# run multiple imputation
tr_mi_raw_mam <- mice(tr_mam, m = 25, maxit = 100, seed = 20)
# method = pmm predictive mean matching

# summary of multiple imputation results including predictor matrix (which variables were used to predict missing values)
summary(tr_mi_raw_mam)

# save: tr_mi_raw_mam
saveRDS(tr_mi_raw_mam, "data/df_tr_mi_raw_mam.rds")
# dataframe of raw multiple imputated trait data for mammals (data.frame)

tr_mi_mam <- lapply(1:25, function(x) {
  out <- mice::complete(tr_mi_raw_mam, action = x) %>% 
    dplyr::select(binomial:diet_pc2) %>% 
    # remove extinct species
    filter(binomial != "Melomys rubicola") %>% 
    filter(binomial != "Pipistrellus murrayi")
})

# save: tr_mi_mam
saveRDS(tr_mi_mam, "data/df_tr_mi_mam.rds")
# list of multiple imputed dataframes for mammals

```

#### Bird trait imputation

```{r}
tr_birds <- filter(tr, class == "Aves") %>% 
  dplyr::select(-class)

## phylogenetic data

# load: t_pem_birds
t_pem_birds <- readRDS("data/df_t_pem_birds.rds")
# phylogenetic eigenvectors matched to species names - using Prum et al., 2015 "A comprehensive phylogeny of birds (Aves) using targeted next-generation DNA sequencing"

# join trait and phylogenetic data (use first ten eigenvectors - Penone et al., 2014)
tr_birds <- left_join(tr_birds, t_pem_birds, by = "binomial")

# missing data pattern
mice::md.pattern(dplyr::select(tr_birds, -binomial, -c(V_1:V_10)))

# run multiple imputation
tr_mi_raw_birds <- mice(tr_birds, m = 25, maxit = 100, seed = 20)
# method = pmm predictive mean matching

# summary of multiple imputation results including predictor matrix (which variables were used to predict missing values)
summary(tr_mi_raw_birds)

# save: tr_mi_raw_birds
saveRDS(tr_mi_raw_birds, "data/df_tr_mi_raw_birds.rds")
# dataframe of raw multiple imputated trait data for birds (data.frame)

tr_mi_birds <- lapply(1:25, function(x) {
  out <- mice::complete(tr_mi_raw_birds, action = x) %>% 
    dplyr::select(binomial:diet_pc2) %>% 
    # remove extinct species
    filter(binomial != "Acrocephalus luscinius") %>% 
    filter(binomial != "Zosterops conspicillatus")
})

# save: tr_mi_birds
saveRDS(tr_mi_birds, "data/df_tr_mi_birds.rds")
# list of multiple imputed dataframes for birds
```

#### Combine mammal and bird complete trait data

```{r}
# combine mammals and birds
tr_mi <- lapply(1:25, function(x) {
  bind_rows(tr_mi_mam[x], tr_mi_birds[x]) %>% 
    arrange(binomial)
})
  
# save: tr_mi
saveRDS(tr_mi, "data/df_tr_mi.rds")
# list of multiple imputed dataframes
```

#### --------------------------------------------------------------
## Step 4: PCA ##
#### --------------------------------------------------------------

```{r}
#### z-transformation ####

# function to z-transform data
scale_z <- function(x){
  (x - mean(x, na.rm=TRUE))/sd(x, na.rm=TRUE)
}

## tr_mi_z ##

# z-transform trait data
tr_mi_z <- lapply(1:25, function(x) {
  tr_mi[[x]] %>%
    # scale (zero mean and unit variance) all traits
    dplyr::mutate_at(vars(body_mass_median:diet_pc2), funs(scale_z)) %>% 
    # convert to true dataframe
    as.data.frame
})

tr_mi_z_d1 <- lapply(1:25, function(x) {
  tr_mi_z[[x]] %>% 
    # add species names to rownames
    tibble::column_to_rownames(var = "binomial") 
})

# run pca
prin_mi_d1 <- lapply(1:25, function(x) {
  princomp((tr_mi_z_d1[[x]]), cor = TRUE, scores = TRUE)
})

comp_var <- lapply(1:25, function(x) {prin_mi_d1[[x]]$sdev^2}) %>% 
  dplyr::bind_cols() %>% 
  dplyr::mutate_all(as.numeric) %>% 
  dplyr::mutate(mean = apply(.[1:25], 1, mean))

comp_var$mean/sum(comp_var$mean)

comp_var$V2/sum(comp_var$V2)

pc_mi_d1 <- lapply(1:25, function(x) {
  # pca scores
  as.data.frame(prin_mi_d1[[x]]$scores) %>% 
    tibble::rownames_to_column("binomial") %>% 
    # add identifier for each imputation dataset
    dplyr::mutate(., mi = paste0("mi_", x))
  }) %>% 
  dplyr::bind_rows()

# function to rescale data from 0-1
rescale3 <- function(x){(x-min(x, na.rm = TRUE))/(max(x, na.rm = TRUE)-min(x, na.rm = TRUE))}

pc_mi_d1 <- pc_mi_d1 %>% 
  # convert long to wide
  tidyr::gather(key, value, -binomial, -mi) %>% 
  tidyr::unite(col, key, mi) %>% 
  tidyr::spread(col, value) %>% 
  # calculate mean per species for principal components
  dplyr::mutate(Comp.1_mean = apply(dplyr::select(., contains("Comp.1")), 1, mean)) %>% 
  dplyr::mutate(Comp.2_mean = apply(dplyr::select(., contains("Comp.2")), 1, mean)) %>% 
  dplyr::mutate(Comp.3_mean = apply(dplyr::select(., contains("Comp.3")), 1, mean)) %>% 
  dplyr::mutate(Comp.4_mean = apply(dplyr::select(., contains("Comp.4")), 1, mean)) %>% 
  dplyr::mutate(Comp.5_mean = apply(dplyr::select(., contains("Comp.5")), 1, mean))

# loadings
pcload_mi_d1 <- lapply(1:25, function(x) {
  # extract pca loadings
  as.data.frame(unclass(prin_mi_d1[[x]]$loadings)) %>% 
    tibble::rownames_to_column("trait") %>% 
    # add identifier for each imputation dataset
    dplyr::mutate(., mi = paste0("mi_", x))
}) %>% 
  dplyr::bind_rows()

pcload_mi_d1 <- pcload_mi_d1 %>% 
  # convert long to wide
  tidyr::gather(key, value, -trait, -mi) %>% 
  tidyr::unite(col, key, mi) %>% 
  tidyr::spread(col, value) %>% 
  # calculate mean per species for principal components
  dplyr::mutate(Comp.1_mean = apply(dplyr::select(., contains("Comp.1")), 1, mean)) %>% 
  dplyr::mutate(Comp.2_mean = apply(dplyr::select(., contains("Comp.2")), 1, mean)) %>% 
  dplyr::mutate(Comp.3_mean = apply(dplyr::select(., contains("Comp.3")), 1, mean)) %>% 
  dplyr::mutate(Comp.4_mean = apply(dplyr::select(., contains("Comp.4")), 1, mean)) %>% 
  dplyr::mutate(Comp.5_mean = apply(dplyr::select(., contains("Comp.5")), 1, mean))

# scalar to adjust arrow size
sc_mi <- 7

pcload_mi_d1_sc <- pcload_mi_d1 %>% 
  # rescale for arrow sizes
  dplyr::mutate_at(vars(contains("Comp")), funs(.*sc_mi)) %>% 
  # posh names
  mutate(trait = c("Body mass", "Diet", "Generation length", "Habitat breadth", "Litter/clutch size"))

# kernel density estimation
pc_raw_mi_d1 <- pc_mi_d1 %>% 
  # extract first two principal components
  dplyr::select(., binomial, Comp.1_mean, Comp.2_mean) %>% 
  tibble::column_to_rownames(var = "binomial")

# optimal bandwidth estimation
hpi_mi_d1 <- Hpi(x = pc_raw_mi_d1)

# kernel density estimation    
est_mi_d1 <- kde(x = pc_raw_mi_d1, H = hpi_mi_d1, compute.cont = TRUE)  

den_mi_d1 <- list(est_mi_d1$eval.points[[1]], est_mi_d1$eval.points[[2]], est_mi_d1$estimate)
names(den_mi_d1) <- c("x", "y", "z")
dimnames(den_mi_d1$z) <- list(den_mi_d1$x, den_mi_d1$y)
dcc_mi_d1 <- melt(den_mi_d1$z)

# kernel function

cl <- function(df, prob) {
  dx <- diff(df$x[1:2])
  dy <- diff(df$y[1:2])
  sz <- sort(df$z)
  c1 <- cumsum(sz) * dx * dy
  approx(c1, sz, xout = 1 - prob)$y
}

# 0.5 probability kernel
cl_50_mi_d1 <- cl(df = den_mi_d1, prob = 0.50)
# 0.95 probability kernel
cl_95_mi_d1 <- cl(df = den_mi_d1, prob = 0.95)
# 0.99 probability kernel
cl_99_mi_d1 <- cl(df = den_mi_d1, prob = 0.99)

# save principal component data
write.csv(pc_mi_d1, file = "data/pc.csv", row.names = FALSE)
```

#### PCA plot

```{r}
# colour palette
col_pal <- colorRampPalette(c("red", "yellow", "white"))(200)

# plot
pca_plot_mi_d1 <- ggplot(dcc_mi_d1, aes(x = Var1, y = Var2)) +
  # coloured probabilty background
  geom_raster(aes(fill = value)) +
  scale_fill_gradientn(colours = rev(col_pal)) +
  # points for species
  geom_point(data = pc_mi_d1, aes(x = Comp.1_mean, y = Comp.2_mean), size = 0.3, alpha = 0.5, colour = "grey20") +
  # probability kernels
  geom_contour(aes(z = value), breaks = cl_50_mi_d1, colour = "grey30", size = 1) +
  geom_contour(aes(z = value), breaks = cl_95_mi_d1, colour = "grey60", size = 1) +
  geom_contour(aes(z = value), breaks = cl_99_mi_d1, colour = "grey70", size = 1) +
  coord_equal() +
  # add arrows
  geom_segment(data = pcload_mi_d1_sc, aes(x = 0, y = 0, xend = Comp.1_mean, yend = Comp.2_mean), arrow = arrow(length = unit(0.2, "cm")), colour = "black") +
  # add dashed arrows ends
  geom_segment(data = pcload_mi_d1_sc, aes(x = 0, y = 0, xend = -Comp.1_mean, yend = -Comp.2_mean), lty = 5, colour = "darkgrey") +
  # add arrow labels
  geom_text(data = pcload_mi_d1_sc, aes(x = Comp.1_mean, y = Comp.2_mean, label = trait), size = 6, nudge_x = c(0, 0, 0, 0, -0.2), nudge_y = c(0.2, -0.2, -0.2, -0.2, 0.2)) +
  # axis labels - see comp_var
  labs(x = "PC1 (34.1%)", y = "PC2 (26.0%)") +
  # edit plot
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text = element_text(colour = "black"),
        legend.position = "none",
        text = element_text(size = 20))

# display plot
pca_plot_mi_d1
```

#### --------------------------------------------------------------
## Step 5: hypervolumes ##
#### --------------------------------------------------------------

#### Use single randomly selected trait imputation dataset

```{r}
# z-transform trait data
tr_mi_z_d1 <- tr_mi[[2]] %>%
  dplyr::mutate_at(vars(body_mass_median:diet_pc1), funs(scale_z)) %>% 
  # scale (zero mean and unit variance) all traits
  as.data.frame %>% 
  # convert to true dataframe ato prevent 'Warning message: Setting row names on a tibble is deprecated.'
  tibble::column_to_rownames(var = "binomial") %>% 
  # add species names to rownames
  dplyr::select(-diet_pc2)
```

#### Observed hypervolume

```{r}
# calculate observed hypervolume
#set.seed(3
#obs_hyper_full_mi <- hypervolume::hypervolume_svm(tr_mi_z_d1)

# save: obs_hyper_full_mi
#saveRDS(obs_hyper_full_mi, "data/hv_obs_hyper.rds")

# load: obs_hyper_full_mi
obs_hyper_full_mi <- readRDS("data/hv_obs_hyper.rds")

# observed volume
obs_hyper_mi <- obs_hyper_full_mi@Volume
```

#### Number of permutations for nulls

```{r}
# number of permutations
npermute <- 999
```

#### Null model 1 (see Diaz et al., 2016)

```{r}
# set up null trait data
nullt1_mi <- replicate(npermute, apply(tr_mi_z_d1, 2, function(x) runif(nrow(tr_mi_z_d1), min = min(x), max = max(x))), simplify = FALSE)

# simulate null hypervolumes
sim_hyper1_mi <- lapply(nullt1_mi, hypervolume_svm)

# 39 secs per hypervolume (11 hours for 999 hypervolumes)

# save: sim_hyper1_mi
saveRDS(sim_hyper1_mi, "data/df_sim_hyper1_mi.rds")

# load: sim_hyper1_mi (not included on Github as the file is too big)
#sim_hyper1_mi <- readRDS("data/df_sim_hyper1_mi.rds")

# null hypervolume volumes
sim_vol1_mi <- sapply(sim_hyper1_mi, function(x) return(x@Volume))

# Permutation test between observed hypervolume and simulated hypervolumes
rtest_hyper1_mi <- ade4::as.randtest(obs = obs_hyper_mi, sim = sim_vol1_mi, alter = "less")
rtest_hyper1_mi
```

#### Null model 2 (see Diaz et al., 2016)

```{r}
# set up null trait data
nullt2_mi <- replicate(npermute, scale(matrix(rnorm(ncol(tr_mi_z_d1) * nrow(tr_mi_z_d1)), nrow(tr_mi_z_d1), ncol(tr_mi_z_d1))), simplify = FALSE)

# add column names for hypervolume function
nullt2_mi <- lapply(nullt2_mi, function(x) {colnames(x) <- colnames(tr_mi_z_d1); x})

# simulate null hypervolumes
sim_hyper2_mi <- lapply(nullt2_mi, hypervolume_svm)

# 145 secs per hypervolume (40 hours for 999)

# save: sim_hyper2_mi
saveRDS(sim_hyper2_mi, "data/df_sim_hyper2_mi.rds")

# load: sim_hyper2_mi (not included on Github as the file is too big)
#sim_hyper2_mi <- readRDS("data/df_sim_hyper2_mi.rds")

# null hypervolume volumes
sim_vol2_mi <- sapply(sim_hyper2_mi, function(x) return(x@Volume))

# Permutation test between observed hypervolume and simulated hypervolumes
rtest_hyper2_mi <- ade4::as.randtest(obs = obs_hyper_mi, sim = sim_vol2_mi, alter = "less")
rtest_hyper2_mi
```

#### Null model 3 (see Diaz et al., 2016)

```{r}
# set up null trait data
nullt3_mi <- replicate(npermute, apply(tr_mi_z_d1, 2, sample), simplify = FALSE)

# simulate null hypervolumes
sim_hyper3_mi <- lapply(nullt3_mi, hypervolume_svm)

# 119 secs (33 hours)

# save: sim_hyper3_mi
saveRDS(sim_hyper3_mi, "data/df_sim_hyper3_mi.rds")

# load: sim_hyper3_mi (not included on Github as the file is too big)
#sim_hyper3_mi <- readRDS("data/df_sim_hyper3_mi.rds")

# null hypervolume volumes
sim_vol3_mi <- sapply(sim_hyper3_mi, function(x) return(x@Volume))

# Permutation test between observed hypervolume and simulated hypervolumes
rtest_hyper3_mi <- ade4::as.randtest(obs = obs_hyper_mi, sim = sim_vol3_mi, alter = "less")
rtest_hyper3_mi
```

#### Null model 4 (see Diaz et al., 2016)

```{r}
# correlation matrix
corM <- cor(tr_mi_z_d1)

# set up null trait data
nullt4_mi <- replicate(npermute, scale(matrix(rnorm(ncol(tr_mi_z_d1) * nrow(tr_mi_z_d1)), nrow(tr_mi_z_d1), ncol(tr_mi_z_d1))), simplify = FALSE)
nullt4_mi <- lapply(nullt4_mi, function(x) scale(x %*% chol(corM)))

# simulate null hypervolumes
sim_hyper4_mi <- lapply(nullt4_mi, hypervolume_svm)

# 62 secs per hypervolme (17 hours for 999)

# save: sim_hyper4_mi
saveRDS(sim_hyper4_mi, "data/df_sim_hyper4_mi.rds")

# load: sim_hyper4_mi (not included on Github as the file is too big)
#sim_hyper4_mi <- readRDS("data/df_sim_hyper4_mi.rds")

# null hypervolume volumes
sim_vol4_mi <- sapply(sim_hyper4_mi, function(x) return(x@Volume))

# Permutation test between observed hypervolume and simulated hypervolumes
rtest_hyper4_mi <- ade4::as.randtest(obs = obs_hyper_mi, sim = sim_vol4_mi, alter = "less")
rtest_hyper4_mi
```

#### Ratio between observed volume and null models

```{r}
# null model 1
mean(obs_hyper_mi/sim_vol1_mi) * 100

# null model 2
mean(obs_hyper_mi/sim_vol2_mi) * 100

# null model 3
mean(obs_hyper_mi/sim_vol3_mi) * 100

# null model 4
mean(obs_hyper_mi/sim_vol4_mi) * 100
```

#### Birds vs mammals (hypervolume overlap)

```{r}
# dataframe of trait data for birds and mammals
tr_mi_z_d1_bm <- tr_mi_z_d1 %>% 
  tibble::rownames_to_column(var = "binomial") %>% 
  dplyr::left_join(bi_class, by = "binomial")

# birds trait data
tr_mi_z_d1_b <- tr_mi_z_d1_bm %>% 
  dplyr::filter(class == "Aves") %>% 
  dplyr::select(-class) %>% 
  tibble::column_to_rownames("binomial")

# mammals trait data
tr_mi_z_d1_m <- tr_mi_z_d1_bm %>% 
  dplyr::filter(class == "Mammalia") %>% 
  dplyr::select(-class) %>% 
  tibble::column_to_rownames("binomial")

# birds hypervolume
#set.seed(3)
#birds_hyper_mi <- hypervolume::hypervolume_svm(tr_mi_z_d1_b, name = "birds")

# save: birds_hyper_mi
#saveRDS(birds_hyper_mi, "data/hv_birds_hyper_mi.rds")

# load: birds_hyper_mi
birds_hyper_mi <- readRDS("data/hv_birds_hyper_mi.rds")

# mammals hypervolume
#set.seed(3)
#mams_hyper_mi <- hypervolume::hypervolume_svm(tr_mi_z_d1_m, name = "mammals")

# save: mams_hyper_mi
#saveRDS(mams_hyper_mi, "data/hv_mams_hyper_mi.rds")

# load: mams_hyper_mi
mams_hyper_mi <- readRDS("data/hv_mams_hyper_mi.rds")

# set mammals and birds hypervolumes
set.seed(3)
bm_set_mi <- hypervolume::hypervolume_set(birds_hyper_mi, mams_hyper_mi, check.memory = FALSE)

# overlap statistics
bm_over_mi <- hypervolume::hypervolume_overlap_statistics(bm_set_mi)

# summarise volumes
hypervolume::get_volume(bm_set_mi)
```

# Plot 3D hypervolumes for mammals and birds

```{r}
bm_set_mi@HVList$HV1 <- NULL
bm_set_mi@HVList$HV2 <- NULL
bm_set_mi@HVList$Union <- NULL

# trait names
names <- c("Body mass", "Litter/clutch size", "Generation length", "Habitat breadth", "Diet")

colnames(bm_set_mi@HVList$Intersection@RandomPoints) <- names
colnames(bm_set_mi@HVList$Unique_1@RandomPoints) <- names
colnames(bm_set_mi@HVList$Unique_2@RandomPoints) <- names

colnames(bm_set_mi@HVList$Intersection@RandomPoints)

# plot hypervolumes for mammals and birds
hypervolume::plot.HypervolumeList(bm_set_mi, show.3d = TRUE, plot.3d.axes.id = c(1,5,3), show.random = FALSE, show.data = TRUE, show.centroid = FALSE, show.density = FALSE, cex.random = 5, colors = c("#C77CFF", "#F8766D","#00BFC4"))
```

#### --------------------------------------------------------------
## Step 6: extinction scenario ##
#### --------------------------------------------------------------

#### Perform probabilistic extinction scenarios

```{r}
tr_mi_z_d1 <- tr_mi_z_d1 %>% 
  tibble::rownames_to_column(var = "binomial")

# add extinction risk data
tr_ex <- left_join(tr_mi_z_d1, iucn_ex, by = "binomial") %>% 
  # DD species treated as LC
  mutate(ex = ifelse(ex == "DD", "LC", ex)) %>% 
  # EW species treated as CR
  mutate(ex = ifelse(ex == "EW", "CR", ex))

tr_ex$ex = factor(tr_ex$ex, levels = c("LC", "NT", "VU", "EN", "CR"))

# sample data to predict extinctions
prob_ex_m100 <- replicate(npermute, tr_ex %>% 
                            # extinction probabilities - moeers et al., 2008 
                            # 100 years into the future
                            dplyr::mutate(prob = dplyr::recode(ex, 
                                                               LC = 0.0001,
                                                               NT = 0.01,
                                                               VU = 0.1,
                                                               EN = 0.667,
                                                               CR = 0.999)) %>% 
                            # group by IUCN category
                            dplyr::group_by(ex) %>%
                            # nest columns within dataframe
                            tidyr::nest() %>%
                            # sample frac per IUCN category to probability of extinctions
                            dplyr::mutate(samp = purrr::map(data, ~sample_frac(., unique(.$prob)))) %>% 
                            # select data
                            dplyr::select(ex, samp) %>% 
                            # unnest - turn data back into dataframe
                            tidyr::unnest(cols = c(samp)), simplify = FALSE)

table(prob_ex_m100[[1]]$ex)

## Extinction risk scenario

## Mooers 100 years - 1,095 spp
tr_ex_m100 <- lapply(1:npermute, function(x) {
  tr_ex %>% 
  # remove species predicted to go extinct
  dplyr::filter(!(binomial %in% prob_ex_m100[[x]]$binomial)) %>% 
  dplyr::select(-ex) %>% 
  tibble::column_to_rownames(var = "binomial")
})

hyper_ex_m100 <- lapply(tr_ex_m100, hypervolume_svm)

# 57 secs per hypervolume (~16 hours for 999)

# save: hyper_ex_m100
saveRDS(hyper_ex_m100, "data/hv_ex_m100.rds")

# load: hyper_ex_m100 (not included on Github as the file is too big)
#hyper_ex_m100 <- readRDS("data/hv_ex_m100.rds")

exp_vol_m100 <- sapply(hyper_ex_m100, function(x) return(x@Volume)) %>% 
  as.data.frame()

colnames(exp_vol_m100) <- c("vol")

# null

null_samp <- replicate(npermute, dplyr::sample_n(tr_ex, nrow(prob_ex_m100[[1]])), simplify = FALSE)

null <- lapply(null_samp, function(x) {
  a <- dplyr::filter(tr_ex, !binomial %in% x$binomial)
  b <- a %>% 
    dplyr::select(-ex) %>% 
    tibble::column_to_rownames(var = "binomial")})

null_hyper <- lapply(null, hypervolume_svm)

# 47 secs per hypervolume (~13 hours for 999)

# save: null_hyper
saveRDS(null_hyper, "data/hv_null_m100_hyper.rds")

# load: null_hyper (not included on Github as the file is too big)
#null_hyper <- readRDS("data/hv_null_m100_hyper.rds")

null_vol <- sapply(null_hyper[1:npermute], function(x) return(x@Volume)) %>% 
  as.data.frame()

colnames(null_vol) <- c("vol")

# compare predicted to null

# kololmogorov-smirnov test
ks.test(exp_vol_m100$vol, null_vol$vol, alternative = "greater") # greater - distributions for which x is stochastically smaller than y (the CDF of x lies above and hence to the left of that for y)

# effect size
obs_hyper@Volume - mean(exp_vol_m100$vol)

obs_hyper@Volume - mean(null_vol$vol)

diff <- data.frame(obs = rep(obs_hyper@Volume, npermute), exp = exp_vol_m100$vol, null = null_vol$vol) %>% 
  dplyr::mutate(exp_diff = obs - exp) %>% 
  dplyr::mutate(null_diff = obs - null)

quantile(diff$exp_diff, probs = c(0.025, 0.975))

quantile(diff$null_diff, probs = c(0.025, 0.975))

```

#### Plot difference between projected and randomized scenarios

```{r}
group <- c(rep("exp", npermute), rep("null", npermute))
dat <- data.frame(vol = c(exp_vol_m100$vol, null_vol$vol), group = group)

desc <- with(dat, psych::describeBy(x = vol, group = group))

nam <- names(desc)
mean <- c(desc$exp$mean, desc$null$mean)
se <- c(desc$exp$se, desc$null$se)
pd <- data.frame(nam, mean, se)

# reorder scenarios
pd$nam <- reorder(pd$nam, -pd$mean)

dp <- ggplot(pd, aes(x = nam, y = mean, colour = nam)) +
  # plot empty data - ensures scenarios stay in order
  geom_point(colour = "white") +
  # add line for observed extant volume
  geom_hline(yintercept = obs_hyper@Volume, lty = 2, lwd = 1) +  
  # add jittered points
  geom_jitter(data = dat, aes(x = group, y = vol, colour = group), alpha = 0.2, width = 0.25) +
  # add violin of data density
  geom_violin(data = dat, aes(x = group, y = vol, colour = group), fill = NA, lwd = 0.5, width = 0.5) +
  # add mean and CI
  geom_crossbar(data = pd, aes(ymax = mean, ymin = mean), width = 0.4) +
  # rename x tick marks
  scale_x_discrete(labels = c("obs" = "Extant", "null" = "Randomized extinction", "exp" = "Projected extinction")) +
  # manually select colours
  scale_colour_manual(values = c("#FF7F45", "darkgrey", "black")) +
  # labels
  labs(x = "", y = bquote("Volume "~(SD^5))) +
  # edit graph
  theme(legend.position = "none")

dp
```

#### Trait permutation tests

```{r}
## set up data ##

# observed
tr_ex_td <- dplyr::select(tr_mi, body_mass_median:diet_pc1) %>% 
  mutate(run = as.character(1))

# expected
tr_ex_m100_td_prep <- lapply(1:npermute, function(x) {
  tr_mi %>% 
    # remove species predicted to go extinct
    filter(!(binomial %in% prob_ex_m100[[x]]$binomial)) %>% 
    column_to_rownames(var = "binomial")
})

tr_ex_m100_td <- bind_rows(tr_ex_m100_td_prep, .id = "run")

# null
# same extinctions for non z-transformed data
null_td <- lapply(null_samp, function(x) {
  a <- filter(tr_mi, !binomial %in% x$binomial)
  b <- a %>% 
    column_to_rownames(var = "binomial")})

tr_null_td <- bind_rows(null_td, .id = "run")

td <- bind_rows(tr_ex_td, tr_ex_m100_td, .id = "df") %>% 
  mutate(df_run = paste0(df, "_", run))

## body mass ##

# mean body mass across runs
bm_mean <- tr_ex_m100_td %>% 
  group_by(run) %>% 
  summarise_at(vars(body_mass_median), funs(mean)) %>% 
  mutate(body_mass_median = 10^body_mass_median) %>% 
  mutate(red = 100 - 10^mean(tr_ex_td$body_mass_median) / body_mass_median * 100)

# compare observed mean body mass to predicted mean body mass (999 runs)
tbm <- as.randtest(obs = 10^mean(tr_ex_td$body_mass_median), sim = bm_mean$body_mass_median, alter = "two-sided")
tbm
plot(tbm, main = "", xlab = "body mass (g)")

## generation length ##

# mean generation length across runs
gl_mean <- tr_ex_m100_td %>% 
  group_by(run) %>% 
  summarise_at(vars(GL), funs(mean)) %>% 
  mutate(GL = 10^GL) %>% 
  mutate(red = 100 - 10^mean(tr_ex_td$GL) / GL * 100)

# compare observed mean GL to predicted mean GL (999 runs)
tgl <- as.randtest(obs = 10^mean(tr_ex_td$GL), sim = gl_mean$GL, alter = "two-sided")
tgl
plot(tgl, main = "", xlab = "generation length (years)")

## litter/clutch size ##

# mean litter/clutch size across runs
ls_mean <- tr_ex_m100_td %>% 
  group_by(run) %>% 
  summarise_at(vars(litter_clutch_size), funs(mean)) %>% 
  mutate(litter_clutch_size = 10^litter_clutch_size) %>% 
  mutate(red = 100 - 10^mean(tr_ex_td$litter_clutch_size) / litter_clutch_size * 100)

# compare observed mean litter/clutch size to predicted mean litter/clutch size
tls <- as.randtest(obs = 10^mean(tr_ex_td$litter_clutch_size), sim = ls_mean$litter_clutch_size, alter = "two-sided")
tls
plot(tls, main = "", xlab = "litter/clutch size")

## habitat breadth ##

# mean habitat breadth across runs
hb_mean <- tr_ex_m100_td %>% 
  group_by(run) %>% 
  summarise_at(vars(hab_breadth), funs(mean)) %>% 
  mutate(hab_breadth = hab_breadth^2) %>% 
  mutate(red = 100 - mean(tr_ex_td$hab_breadth)^2 / hab_breadth * 100)

# compare observed mean habitat breadth to predicted mean habitat breadth
thb <- as.randtest(obs = mean(tr_ex_td$hab_breadth)^2, sim = hb_mean$hab_breadth, alter = "two-sided")
thb
plot(thb, main = "", xlab = "habitat breadth")

## diet ##

# mean diet across runs
di_mean <- tr_ex_m100_td %>% 
  group_by(run) %>% 
  summarise_at(vars(diet_pc1), funs(mean)) %>% 
  mutate(diet_pc1 = diet_pc1) %>% 
  mutate(red = 100 - mean(tr_ex_td$diet_pc1) / diet_pc1 * 100)

# compare observed mean diet to predicted mean diet
tdi <- as.randtest(obs = mean(tr_ex_td$diet_pc1), sim = di_mean$diet_pc1, alter = "two-sided")
tdi
plot(tdi, main = "", xlab = "diet")

# plot all permutation plots together
png(filename = "outputs/perm_plots.png", width = 14, height = 10, units = "in", res = 300)
par(mfrow = c(2,3))
plot(tbm, main = "", xlab = "Body mass (g)")
plot(tgl, main = "", xlab = "Generation length (years)")
plot(thb, main = "", xlab = "Litter/clutch size")
plot(thb, main = "", xlab = "Habitat breadth")
plot(tdi, main = "", xlab = "Diet")
dev.off()
```

